Verification of Stencil Computations through Spatial Specifications

### 2016-07-03 paper.tex
Spelling corrections

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteratively computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil comptuations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c). Pictoral representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

Our specification system is based on the observation
that most forms of array access in numberical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large). We found that indeed \dnote{..}.

We outline the specification language here. Section~\ref{sec:syntax}
oultines the syntax. Section~\ref{sec:semantics} defines its semantics
via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
equational theory for specifications via relation ($\equiv$) and a
theory of approximation via relation ($<:$). These equations are then
proven sound with respect to the set semantics of Section~\ref{sec:semantics}.

@@

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop\footnote{Or equivalent for the target language, \eg{},
    \texttt{do} in Fortan for our implementation}, incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop, and such variables will be ranged
over by $i$, $j$, $k$. Since we do not use non-base induction
variables (which are traditionally affine transformations of induction
variables) we often simply say \emph{induction variable}.
\end{definition}

@@

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\end{definition}

@@

Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
which we introduces in stages below.  The top-level is given by the
\nonterm{specification} production which splits into either a
\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
(specification declaration). Specification declarations associate a
specification to one or more program variables $\bar{v}$ (here given
using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$),
describing how the array variables $\bar{v}$ are acessed.
%Specification declarations are either \nonterm{spatial} or
%\nonterm{temporal} specifications and describe, for spatial
%specifcations,
%Specifications declarations describe
%how an array variable $v$ is read
%, and for temporal
%specifications what variables are used to define the array variable $v$.
%%

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by both $r$ and
$r'$. It can be thought of as a conjunction of speicifcations. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)
         + centered(depth=1, dim=2) :: a

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the speicifcation from the previous
 example can be restated as:
%%
\begin{ExmVerbatim}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= stencil r1*r2 :: a
\end{ExmVerbatim}
%%
\paragraph{Spatial modifiers}
%%
There are three kinds of modifier that can be given to spatial
specifications: \texttt{readOnce}, \texttt{reflexive},
\texttt{irreflexive}.

The \texttt{readOnce} modifier enforces that no index appears
more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{ExmVerbatim}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
However, this would be an invalid specification if any of the
array subscripts was reapeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts.
The notion correponds to that of \emph{linearity}, though we
opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

The \texttt{irreflexive} modifier is parameteresed by a
comma-separated list of dimensions, indicating that for those
dimensions there is no neighbour index with $0$ offset, \eg{},
%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1) + a(i-1)
\end{ExmVerbatim}
%
or with just a \texttt{forward} region an example would be:
%
\begin{ExmVerbatim}
!= region r = forward(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1)
\end{ExmVerbatim}

@@

\paragraph{Equations}
Figure~\ref{fig:equations} lists the equational theory for
specifications. This is broken down into equations on regions only,
given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and on
on specification declarations assigned to variables via the equivalence relation
$\textit{specDec} :: \texttt{v} \equiv
\textit{specDec'} :: \texttt{v}$.

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-determinstically picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise pemutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
\{\infty\}$. In pairwise permutation $\infty$

@@

%The main interpretation function is overloaded on lists of
%variable-spec pairs $\interp{\overline{v : S}}$,
%returning multisets of variable-relative-index pairs. This provides
%the top-level definition of the model, with $\interp{-}$ overloaded
%on $S$, $\interp{S}$ mqpping to multisets of
%relative indices not associated to an array avariable.

@@

\mnote{Few cases that we might like to consider (there are examples
    of each in \#camfort-main Slack channel):
  \begin{itemize}
    \item Matrix operator overloading $a + b$, where both are arrays
    \item Scalar operator overloading $a + b$, where a is an array b is a
      scalar. $b$ is added to all.
    \item Sliding. $a = b$ where the dimensionsare $a(1:10)$, $b(2:11)$,
      so the assignment causes shifting.
  \end{itemize}
}

@@

\subsection{Inference}

We demonstate the main steps of the inference procedure over the
following example program which computes the mean value
of a five-point stencil at each index of the input array:
%%
\begin{ExmVerbatim}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{ExmVerbatim}
%%
\paragraph{Step 1: Standard control and dataflow analyses}
The inference relies on some standard program analyses, computed
before the main inference procedure:
%
\begin{enumerate}
\item basic blocks (CFG);
\item induction variables per basic block;
\item (interprocedural) dataflow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards dataflow to this expression, based on the transitive
  closure of the flows graph);
\item type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.
\end{enumerate}
%

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statment. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
index indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed from each array variable
in this flow set to a representation of the subscripts, computed
with \textsf{convert}.

@@

Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
$n$ whose values are drawn from $\mathbb{Z}_\top$.
We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
indexed) and $\vect{x}_{i:j}$ for the subvector from entry $i$ to
entry $j$ (inclusive).

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box (\emph{hyperrectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vecvtors.
\end{definition}

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional spaace. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be unpermuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

>>>>

\emph{Stencils} are a ubiquitous programming pattern, common in
scientific and numerical computing applications. Informally, a stencil
computation computes an array value, where the value at each index $i$ of
this array is calculated from a \emph{neighbourhood} of values around $i$ in
some input array(s), \eg{}, the Game of Life, convolutions in image
processing, approximations to differential equations. For example, the
following iteration computes the one-dimensional discrete Laplace
transform (an approximation to a derivative) in Fortran:
%
\begin{ExmVerbatim}
do iter = 0, itermax
   do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
   end do
   a = b
end do
\end{ExmVerbatim}
%
Line 3 is the core of the stencil computation, calculating
the value at \texttt{b(i)} from a neighbourhood of elements about
\texttt{a(i)}. Line 4 swaps
\texttt{a} and \texttt{b} between iterations, where \texttt{b} becomes the
input for the next iteration. In Fortran, parentheses \texttt{( )} are used
for array subscripts rather than the more familiar bracket syntax \texttt{[ ]}.

@@

In this work, we introduce a simple specification language for the
spatial properties of stencil computations. The
specifications abstract over the fine grained detail of stencil access
patterns.  In practise, most stencil computations have a regular shape
that can be described simply and abstractly, with a small set of
coarse-grained descriptions. In the case of our first Laplace example,
our inference procedure provides the specification:
%
\begin{SpecVerbatim}
!=  stencil centered(depth=1, dim=1) :: a
\end{SpecVerbatim}
%
This explains that \texttt{a} is read from with a
symmetrical stencil pattern (``centered'') to a depth of one in each
direction in its first (and only) dimension.
%The second line explains
%the temporal relationship between \texttt{b} and \texttt{a}: that
%previous time step for \texttt{b} is actually provided by \texttt{a},
%and vice versa. This is explained as a mutual dependence between
%\texttt{a} and \texttt{b}.
In the case of the Navier-Stokes example of
Figure~\ref{ref:navier-stokes-fragment}, its inferred specification is shown
in Figure~\ref{ref:navier-stokes-fragment}(b). The
specification explains that, over the whole fragment, \texttt{u} is
read from with a centered pattern to depth of 1 in both dimensions
(this is known as the \emph{five point stencil}) and \texttt{v}
is read from in a neighbourhood bounded by forward to depth of $1$ in
the first dimension and backward to a depth of $1$ in the second
dimension. Figure~\ref{ref:navier-stokes-fragment}(c) describes these
pictorially.

@@

\begin{center}
\begin{tikzpicture}
\node at (1.25,1.22) {i, j};
\draw (1,1) rectangle (1.5,1.5);
\draw (1.5,1) rectangle (2,1.5);
\draw (1,0.5) rectangle (1.5,1);
\draw (0.5,1) rectangle (1,1.5);
\draw (1,1.5) rectangle (1.5,2);
%
\node at (4.25,1.22) {i, j};
\draw (4,1) rectangle (4.5,1.5);
\draw (4.5,1) rectangle (5,1.5);
\draw (4.5,1.5) rectangle (5,2);
\draw (4,1.5) rectangle (4.5,2);
\end{tikzpicture}
\end{center}
(c). Pictorial representation of the two stencil specifications.
The horizontal dimensions is \text{dim=1} and the vertical is \texttt{dim=2}:
\caption{Fragment of Navier-Stokes fluid simulator and its specification}
\label{ref:navier-stokes-fragment}
\end{figure}

@@

Our specification system is based on the observation
that most forms of array access in numerical code have
a fixed, statically-determined access pattern. For example, the
``\emph{five-point stencil}'' on a two-dimensional array reads from array
indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid
out-of-bounds access at the edges). We revisit this hypothesis
in Section~\ref{sec:evaluation} where the inference of
such regular stencil patterns on a corpus of numerical programs (both
small and large). We found that indeed \dnote{..}.

We outline the specification language here. Section~\ref{sec:syntax}
outlines the syntax. Section~\ref{sec:semantics} defines its semantics
via a simple set interpretation over indices. Section~\ref{sec:eqs} provides an
equational theory for specifications via relation ($\equiv$) and a
theory of approximation via relation ($<:$). These equations are then
proven sound with respect to the set semantics of Section~\ref{sec:semantics}.

@@

\begin{definition}[Base induction variable]
  A variable \textit{v} of integer type is a \emph{base induction
    variable} if it is the control variable of a ``for''
  loop\footnote{Or equivalent for the target language, \eg{},
    \texttt{do} in Fortran for our implementation}, incremented by $1$ per
iteration. The variable is marked as a base induction variable
only within the scope of the loop, and such variables will be ranged
over by $i$, $j$, $k$. Since we do not use non-base induction
variables (which are traditionally affine transformations of induction
variables) we often simply say \emph{induction variable}.
\end{definition}

@@

\begin{definition}[Neighbourhood index and neighbourhood subscripts]
  For an array subscript $v(\bar{e})$, we say that $e \in \bar{e}$
  is a \emph{neighbourhood index} if it is of the form
  $e \equiv i$, $e \equiv i + a$ or $e \equiv i - a$ where $a$ is a
  constant of integer type. That is, a neighbourhood index is a
  relative index defined as a constant translation of a base induction
  variable. The relation $\equiv$ identifies terms up-to commutativity
  of $+$ and the inverse relation of $+$ and $-$ (\eg{},
  $(-b) + i \equiv i - b$).  We classify neighbourhood indices using
  the predicate \neigh{}. An array subscript $v(\bar{e})$ where every
 $e \in \bar{e}$ is a neighbourhood index is called a
 \emph{neighbourhood subscript}.
\end{definition}

@@

Figure~\ref{fig:syntax} gives the syntax of stencil specifications,
which we introduces in stages below.  The top-level is given by the
\nonterm{specification} production which splits into either a
\nonterm{regionDec} (region declaration) or a \nonterm{specDec}
(specification declaration). Specification declarations associate a
specification to one or more program variables $\bar{v}$ (here given
using the Fortran style syntax of $\ldots \; \texttt{::} \bar{v}$),
describing how the array variables $\bar{v}$ are accessed.
%Specification declarations are either \nonterm{spatial} or
%\nonterm{temporal} specifications and describe, for spatial
%specifications,
%Specifications declarations describe
%how an array variable $v$ is read
%, and for temporal
%specifications what variables are used to define the array variable $v$.
%%

@@

The sum of two regions $r \texttt{+} r'$ specifies that an array is
read using the neighbourhood indices described by both $r$ and
$r'$. It can be thought of as a conjunction of specifications. For
example, the following gives the specification for the five-point
stencil which is the sum of two \texttt{centered} regions in each
dimension:
%%
\begin{ExmVerbatim}
!= stencil centered(depth=1, dim=1)
         + centered(depth=1, dim=2) :: a

@@

Region specifications can be assigned to region variables via
a region declaration, which can be used later to form a spatial
specification. For example, the specification from the previous
 example can be restated as:
%%
\begin{ExmVerbatim}
!= region r1 = centered(depth=1, dim=1)
!= region r2 = centered(depth=1, dim=2)
!= stencil r1*r2 :: a
\end{ExmVerbatim}
%%
\paragraph{Spatial modifiers}
%%
There are three kinds of modifier that can be given to spatial
specifications: \texttt{readOnce}, \texttt{reflexive},
\texttt{irreflexive}.

The \texttt{readOnce} modifier enforces that no index appears
more than once. For example, in all of the previous examples the
\texttt{readOnce} modifier could be added, \eg{}
%
\begin{ExmVerbatim}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{ExmVerbatim}
%
However, this would be an invalid specification if any of the
array subscripts was repeated. This modifier provides a way to
 rule out any accidental repetition of array subscripts.
The notion corresponds to that of \emph{linearity}, though we
opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

The \texttt{irreflexive} modifier is parameterised by a
comma-separated list of dimensions, indicating that for those
dimensions there is no neighbour index with $0$ offset, \eg{},
%
\begin{ExmVerbatim}
!= region r = centered(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1) + a(i-1)
\end{ExmVerbatim}
%
or with just a \texttt{forward} region an example would be:
%
\begin{ExmVerbatim}
!= region r = forward(depth=1, dim=1)
!= stencil irreflexive(dim=1), r :: a
b(i) = a(i+1)
\end{ExmVerbatim}

@@

\paragraph{Equations}
Figure~\ref{fig:equations} lists the equational theory for
specifications. This is broken down into equations on regions only,
given by the equivalence relation $\textit{region} \equiv \textit{region}'$ and
on specification declarations assigned to variables via the equivalence relation
$\textit{specDec} :: \texttt{v} \equiv
\textit{specDec'} :: \texttt{v}$.

@@

The pairwise permutation $u \bowtie v$ takes two vectors and builds the matrix
where the rows are all possible $n$-vectors generated by
non-deterministic picking each $i^{th}$ entry from either the
$i^{th}$ entry in $u$ or the $i^{th}$ entry in $j$. For example:
%
\begin{equation*}
\vthreeh{0}{1}{2} \bowtie \vthreeh{3}{4}{5} =
\setlength{\arraycolsep}{0.5em}
\left[\begin{array}{ccc}
0 & 1 & 2 \\
0 & 1 & 5 \\
0 & 4 & 2 \\
0 & 4 & 5 \\
3 & 1 & 2 \\
3 & 1 & 5 \\
3 & 4 & 2 \\
3 & 4 & 5
\end{array}
\right]
\end{equation*}
%
The $2^n$ unique choices for $\bowtie$ on $n$-vectors
corresponds to taking all bit-strings of length $n$ and
selecting from $u$ for $1$ and $v$ for $0$. Thus, we can
defined the pairwise permutation matrix as:
%
\begin{align*}
(u \bowtie v)_{i,j} & = b_{i,j} u_j + \neg b_{i,j} v_j
\end{align*}
%
where $b$ is a logical matrix where $b_{i,j}$ is the $j$-th bit of the integer
$i - 1$. Recall that the vectors are defined over the domain $\mathbb{Z} \cup
\{\infty\}$. In pairwise permutation $\infty$

@@

%The main interpretation function is overloaded on lists of
%variable-spec pairs $\interp{\overline{v : S}}$,
%returning multisets of variable-relative-index pairs. This provides
%the top-level definition of the model, with $\interp{-}$ overloaded
%on $S$, $\interp{S}$ mapping to multisets of
%relative indices not associated to an array variable.

@@

\mnote{Few cases that we might like to consider (there are examples
    of each in \#camfort-main Slack channel):
  \begin{itemize}
    \item Matrix operator overloading $a + b$, where both are arrays
    \item Scalar operator overloading $a + b$, where a is an array b is a
      scalar. $b$ is added to all.
    \item Sliding. $a = b$ where the dimensions are $a(1:10)$, $b(2:11)$,
      so the assignment causes shifting.
  \end{itemize}
}

@@

\subsection{Inference\label{sec:inference}}

We demonstrate the main steps of the inference procedure over the
following example program which computes the mean value
of a five-point stencil at each index of the input array:
%%
\begin{ExmVerbatim}
do i = 1, (n-1)
   do j = 1, (m-1)
      x       = a(i-1, j) + a(i+1, j)
      y       = a(i, j-1) + a(i, j+1)
      b(i, j) = (a(i, j) + x + y) / 5.0
   end do
end do
\end{ExmVerbatim}
%%
\paragraph{Step 1: Standard control and data-flow analyses}
The inference relies on some standard program analyses, computed
before the main inference procedure:
%
\begin{enumerate}
\item basic blocks (CFG);
\item induction variables per basic block;
\item (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\item type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.
\end{enumerate}
%

@@

For each assignment statement whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed which
maps array variables to a set of vectors representing array
subscripts.  This finite map contains all array subscript expressions
which flow to this statement. More formally, a function
$\textsf{analyse}$ is applied to each statement in a program with the
following clause:
%
\begin{align*}
\begin{array}{lr}
\textsf{analyse}(v(\overline{e_1}) = e_2)
 := \qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \wedge \arrayTy(v)  \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{v' \mapsto \{\textsf{convert}(\bar{e})\} \mid v'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(v')\}}
\end{array}
\end{align*}
%
That is, we focus on assignments to an array subscript where each
index indexing expression in $\bar{e}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed from each array variable
in this flow set to a representation of the subscripts, computed
with \textsf{convert}.

@@

Let $\vect{x}, \vect{y}, \vect{z}$ range over the column vectors of size
$n$ whose values are drawn from $\mathbb{Z}_\top$.
We write $\vect{x}_i$ for the $i$-th value of vector $\vect{x}$ (1
indexed) and $\vect{x}_{i:j}$ for the sub-vector from entry $i$ to
entry $j$ (inclusive).

\begin{definition}[Spans]
  A \emph{span} represents an $n$-dimensional box (\emph{hyper-rectangle}) as
  a pair of $n$-dimensional vectors (represented as a $2 \times n$
  matrix) giving the co-ordinates of the lower-bound vertex (first
  column) and the upper-bound vertex (second column). We use the
  notation $[\vect{x}^L \; \vect{x}^U]$ for such a span to mark out
  the lower and upper bound vectors.
\end{definition}

The algorithm to create contiguous spans, covering the space of
indices, then proceeds as follows.
Firstly, for all $v \in \mathsf{dom}(M)$ then a new map $M'$ is
created where each vector is mapped to the trivial \emph{unit span} is
created by pairing a vector with itself:
%
\begin{align*}
N(v) = \{[\vect{x} \;
  \vect{x}] \mid \vect{x} \in M(v)\}
\end{align*}
%
For our example, this yields:
%%
\begin{align*}
N(\texttt{a}) = \stwo{0}{0}{0}{0} \stwo{1}{0}{1}{0} \stwo{-1}{0}{-1}{0} \stwo{0}{1}{0}{1} \stwo{0}{-1}{0}{-1}
\end{align*}
%%
A fixed point is then compute for the following
procedure \textsf{regions} which, for per variable
in the map, coalesces spans into contiguous regions in
the $n$-dimensional space. That is, we compute
 $(\mu \; \textsf{regions}) N$ where \textsf{regions}
is defined as follows for each variable in the map:
%
\begin{enumerate}

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be un-permuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.
